WASP-4 is Accelerating Towards the Earth

### 2020-01-25 paper/ms.tex
update words

\begin{abstract}
  We recently reported that space and ground-based transit observations
  showed the orbital period of WASP-4b to be decreasing.
  Possible explanations included tidal orbital decay, orbital
  precession, and light-travel time effects.
  In this work we present new observations of the radial velocity of
  WASP-4, acquired with Keck-HIRES.
  The new data show that the system is accelerating towards our line
  of sight, at $\dot{\gamma} = X.XX \pm Y.YY\ {\rm m}\,{\rm
  s}^{-1}\,{\rm yr}^{-1}$.
  This implies a period decrease of X.XX milliseconds per year,
  compared to the observed value of Y.YY.
  Combining the radial velocities with new speckle imaging from
  Gemini-Zorro, we find that the WASP-4 system likely has a
  10-200$\,M_{\rm Jup}$ companion with semi-major axis between 10 and
  100$\,$AU.
  As observing baselines continue to extend, long-term tracking of hot
  Jupiter radial velocities will be necessary to distinguish tidal
  orbital decay from the R{\o}mer delay.
\end{abstract}

@@

Since the first discovery of transiting hot Jupiters, the hunt for
tidal orbital decay has received much attention (CITE LIKE 20 PAPERS).
These efforts are gradually beginning to yield fruit: by combining
transit and occultation timing and radial velocity measurements, CITE
Yee et al{.} have shown that the data for the hot Jupiter WASP-12b are
most compatible with orbital decay.

The present study highlights a point that, while obvious, has perhaps
not yet received due attention.  The point is that the hunt for
orbital decay of hot Jupiters will be crippled without long-term
radial velocity monitoring programs.  The reason is simple: when the
timescale of orbital period change far exceeds the observing baseline,
the first deviation from constant periodicity is always quadratic in
the transit number.

There are ample reasons to expect orbital decay to be common (CITE).
However there are reasons to expect the R{\o}mer delay to perhaps be
even {\it more} common (CITE Knutson 14, Bryan 16).  Both manifest to
first order identically in transit times (up to the sign of the
line-of-sight acceleration) -- as a non-zero period derivative.

Specifically, \citet{knutson_friends_2014} showed that X\% of hot
Jupiters show some form of radial acceleration in radial velocity
time-series with baselines of YY years.  Assuming a separation
distribution of (WHATEVER), this implies that ZZ\% of hot Jupiters
will {\it always} show a changing orbital period, over timescale of
say 10 years.
The abundance of giant planets
exterior to Super-Earth sized systems is also high (CITE Bryan 2019),
but this is not so much of a concern to us, as their planetary masses
and therefore tidal timescales are typically expected to be at least
an order of magnitude shorter than hot Jupiters.

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.  We discovered the timing variations using
data from TESS and a decade of ground-based monitoring
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.

Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  With additional
data, they found a lower best-fit decay rate of $\dot{P} = -XX.X \pm
X.X$ milliseconds per year.

A separate study by \citet{baluev_2019} reported
additional archival transit light curves of WASP-4b, most notably from
the TRAPPIST archive, and a select few other observers.
% \citeauthor{baluev_homogeneously_2019} analyzed their newly obtained
% photometry, along with archival light curves that we omitted from our
% analysis due to systematic uncertainties in the absolute time system.
\citeauthor{baluev_2019} found that when they used all
the available TTV data, the need for a quadratic ephemeris was present
``at the high $\sim 5-7$ sigma level''.  However, they pointed out
that if they used lower-precision subsets of the available timing
data, the necessity for the quadratic term decreased. This is not
particularly surprising.
% \citeauthor{baluev_homogeneously_2019} also pointed out that the
% precise transit times reported by \citet{huitson_gemini_2017} were
% quite important in the time-series.  Overall,
% \citeauthor{baluev_homogeneously_2019} did not find the claim of a
% period decrease convincing.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the radial velocities
were sparse and showed a weak ($<2\sigma$) linear trend.  Our new
measurements reveal a line of sight acceleration of $\dot{\gamma} =
-XX {\rm m\,s^{-1}\,yr^{-1}}$.  This translates to an expected period
decrease simply from the light travel time effect of ZZ milliseconds
per year --- about commensurate with what is observed.

@@

WASP-4 was found to be accelerating towards our line of sight at high
confidence,
\begin{equation}
  \dot{v}_{\rm r} =
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our recent observing run, 
$\dot{v}_{\rm r}$ was thought to be about five times smaller, and was
only marginally significant
\citep{knutson_friends_2014,bouma_wasp4b_2019}.

The system's acceleration towards our line of sight causes a decrease
in the apparent orbital period (the R{\o}mer delay).  The period
derivative expected from radial velocities is
\begin{align}
  \dot{P}_{\rm RV} &= \frac{\dot{v}_{\rm r} P}{c} \\
  \dot{P}_{\rm RV} &= -5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P} = - X.X \pm 1.1~{\rm ms}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.
% FIXME
All the
available TTV data show a period decrease of $\approx 8-12$
milliseconds per year~\citep{bouma_wasp-4b_2019,southworth_transit_2019,baluev_homogeneously_2019}.

@@

Given a linear radial velocity trend, we can place lower-limits on the
mass and semi-major axis of additional bodies in the system.  The
constraints on the two parameters are degenerate: $\dot{\gamma}
\propto M_{\rm p} a^{-2}$.  Nonetheless, the Zorro images and contrast
curves can limit the available parameter space by setting an upper
limit on the semimajor axis, and a maximum brightness (and thereby
mass) of any putative companions.

The general procedure we use to derive constraints on possible
companion masses and semi-major axes has been developed by
\citet{wright_linear_trends_2007}, \citet{crepp_trends_2012},
\citet{montet_trends_2014}, \citet{knutson_friends_2014},
\citet{bryan_statistics_2016,bryan_excess_2019}, and others.

@@

\subsection{How many other hot Jupiters will show the R{\o}mer delay?}

@@

@@

We found that WASP-4 is accelerating towards the Earth, at
X.XX m/s/yr.
This seems like the most plausible explanation for the orbital 
period decrease observed in transits~---~not tidal
orbital decay, or apsidal precession.
The most likely parameters for the companion are XXX.
Over half of hot Jupiters have super-Jovian outer companions,
so the R{\o}mer delay will become an increasingly large nuisance in
the hunt for orbital decay as
the observing baselines for hot Jupiters increase.
The requirement to avoid missing out is to have long-term
radial velocity monitoring with the same spectrograph (or else to
calibrate instrumental offsets by observing the same stars simultaneously).

>>>>

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  Potential causes of the period change include tidal orbital decay,
  orbital precession, and light-travel time effects.
  In this work we present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.
  The new data show that the system is accelerating towards the Earth
  at $\dot{\gamma} = X.XX \pm Y.YY\ {\rm m}\,{\rm s}^{-1}\,{\rm
  yr}^{-1}$.
  The corresponding period decrease explains WASP-4's
  changing orbital period as a light-travel time effect.
  Combining the radial velocities with new speckle imaging limits, we
  find that system's acceleration is likely caused by a
  10-200$\,M_{\rm Jup}$ companion with semi-major axis between 10-100$\,$AU.
  Based on statistics from \citet{knutson_friends_2014}, about 1 in 6
  hot Jupiters are expected to show period decreases commensurate with
  WASP-4 due to acceleration by outer companions.
  Continued long-term radial velocity monitoring of hot Jupiters is
  therefore essential to distinguish tidal orbital decay from
  line-of-sight accelerations.
\end{abstract}

@@

The issue of whether and how the orbits of hot Jupiters decay due to
tidal interaction with their host stars has drawn significant
attention over the past 25 years
\citep{rasio_tidal_1996,levrard_falling_2009,matsumura_tidal_2010}.
Numerous observations have also been performed (CITE).  These efforts
are now yielding fruit: by combining transit and occultation timing
with radial velocity measurements, \citet{yee_orbit_2020} have shown
that the data for the hot Jupiter WASP-12b are most compatible with
orbital decay.

The present study highlights a point that, though obvious, has perhaps
not yet received due attention.  The point is that observational
programs aimed at identifying orbital decay in hot Jupiters will be
crippled without long-term radial velocity monitoring programs.  The
reason is simple: when the timescale of orbital period change exceeds
the observing baseline, the first deviation from constant periodicity
is always quadratic in the transit number.

While orbital decay may be common (CITE).  line-of-sight accelerations
could be even {\it more} common (CITE Knutson 14, Bryan 16).  Both
manifest to first order identically in transit times (up to the sign
of the line-of-sight acceleration) -- as a non-zero period derivative.
Specifically, \citet{knutson_friends_2014} showed that X\% of hot
Jupiters show some form of radial acceleration in radial velocity
time-series with baselines of YY years.  Assuming a separation
distribution of (WHATEVER), this implies that ZZ\% of hot Jupiters
will {\it always} show a changing orbital period, over timescale of
say 10 years.

% The abundance of giant planets
% exterior to Super-Earth sized systems is also high (CITE Bryan 2019),
% but this is not so much of a concern to us, as their planetary masses
% and therefore tidal timescales are typically expected to be at least
% an order of magnitude shorter than hot Jupiters.

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.  We discovered the timing variations using
data from TESS and a decade of ground-based monitoring
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.
Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  With additional
data, they found a lower best-fit decay rate of $\dot{P} = -XX.X \pm
X.X$ milliseconds per year.
A separate study by \citet{baluev_2019} reported
additional archival transit light curves of WASP-4b, most notably from
the TRAPPIST archive, and a select few other observers.
% \citeauthor{baluev_homogeneously_2019} analyzed their newly obtained
% photometry, along with archival light curves that we omitted from our
% analysis due to systematic uncertainties in the absolute time system.
% \citeauthor{baluev_2019} found that when they used all the available
% TTV data, the need for a quadratic ephemeris was present.  However,
% they pointed out that if they used lower-precision subsets of the
% available timing data, the necessity for the quadratic term decreased.
% This is not particularly surprising.
% \citeauthor{baluev_homogeneously_2019} also pointed out that the
% precise transit times reported by \citet{huitson_gemini_2017} were
% quite important in the time-series.  Overall,
% \citeauthor{baluev_homogeneously_2019} did not find the claim of a
% period decrease convincing.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the radial velocities
were sparse and showed a weak ($<2\sigma$) linear trend.  Our new
measurements reveal a line-of-sight acceleration of $\dot{\gamma} =
-XX {\rm m\,s^{-1}\,yr^{-1}}$.  This translates to an expected period
decrease simply from the light travel time effect of ZZ milliseconds
per year --- about commensurate with what is observed.
In the following, we avoid the term ``R{\o}mer delay'' as this phrase
typically signifies arrival time delay due to observatory motion. We
are discussing an analog of the Doppler effect for a source with
constant line-of-sight acceleration.

@@

WASP-4 was found to be accelerating towards our line-of-sight at high
confidence,
\begin{equation}
  \dot{v}_{\rm r} =
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our recent observing run, 
$\dot{v}_{\rm r}$ was thought to be about five times smaller, and was
only marginally significant
\citep{knutson_friends_2014,bouma_wasp4b_2019}.

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The period
derivative expected from radial velocities is
\begin{align}
  \dot{P}_{\rm RV} &= \frac{\dot{v}_{\rm r} P}{c} \\
  \dot{P}_{\rm RV} &= -5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P} = - X.X \pm 1.1~{\rm ms}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.
% FIXME
All the
available TTV data show a period decrease of $\approx 8-12$
milliseconds per year~\citep{bouma_wasp-4b_2019,southworth_transit_2019,baluev_homogeneously_2019}.

@@

Given a linear radial velocity trend, we can place lower-limits on the
mass and semi-major axis of additional bodies in the system.  As a
quick estimate of the minimum mass required to explain the linear
trend in WASP-4, we turn to \citet{feng_california_2015}.  As they
discuss, the scenario that yields the minimum companion mass for a
system with a linear trend is a companion with $e\approx0.5$ and
$\omega=90^\circ$.  From their Equation~1,
\begin{equation}
 M_{\rm min} \approx 0.0164\,M_{\rm Jup}
  \left( \frac{\tau}{{\rm yr}} \right)^{4/3}
  \left| \frac{\dot{\gamma}}{{\rm m\,s^{-1}\,yr^{-1}}} \right|
  \left( \frac{M_\star}{{M_\odot}} \right)^{2/3},
\end{equation}
where $\tau$ is the observing baseline.
For WASP-4, this yields $M_{\rm min} = 4.7 M_{\rm Jup}$.  Higher mass
companions are allowed, presuming that they orbit further from the
star; at fixed $\dot{\gamma}$, $M_{\rm comp} \propto a^2$
\citep{torres_substellar_1999,liu_crossing_2002}.

High-resolution images can further limit the available parameter space
by setting an upper limit on the semimajor axis, and a maximum
brightness (and thereby mass) of any putative companions.
The procedure we use to combine constraints from both radial
velocities and high resolution imaging has been developed by
\citet{wright_linear_trends_2007}, \citet{crepp_trends_2012},
\citet{montet_trends_2014}, \citet{knutson_friends_2014},
\citet{bryan_statistics_2016,bryan_excess_2019}, and others.

@@

\subsection{How many other hot Jupiters are accelerating towards the
Earth?}

@@

Including WASP-4b, 16 of 51 hot Jupiters from
\citet{knutson_friends_2014} show a non-zero radial velocity trend.
About half are positive, and half are negative, and therefore
$\approx$1 in 6 hot Jupiters are expected to show period decreases
commensurate with WASP-4 due to acceleration by outer companions.

@@

We found that WASP-4 is accelerating towards the Earth, at
X.XX m/s/yr.
This seems like the most plausible explanation for the orbital 
period decrease observed in transits~---~not tidal
orbital decay, or apsidal precession.
The most likely parameters for the companion are XXX.
Over half of hot Jupiters have super-Jovian outer companions,
so the light-travel time effect will become an increasingly large nuisance in
the hunt for orbital decay as
the observing baselines for hot Jupiters increase.
The requirement to avoid missing out is to have long-term
radial velocity monitoring with the same spectrograph (or else to
calibrate instrumental offsets by observing the same stars simultaneously).
